Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Getting evaluation in working order.

\begin{figure}[t]
\begin{center}
\includegraphics[width=3.2in]{church-relative-space.ps}
\includegraphics[width=3.2in]{church-relative-speed.ps}
\includegraphics[width=3.2in]{church-relative-time.ps}
\vspace{-1.5em}
\end{center}
\caption{Relative improvements over the baseline analyzer for the
  \Church{} benchmark in terms of peak memory usage, the rate of state
  transitions, and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

We have implemented, optimized, and evaluated \emph{all} of the
machines presented in their JFP paper
\cite{dvanhorn:VanHorn2012Systematic}, with the exception of the
machine for performing pushdown flow analysis.  This includes machines
supporting higher-order functions, state, first-class control,
laziness, and stack inspection.

For each benchmark, we collect analysis times, peak memory usage, and
the rate of states-per-second explored by the analysis for each of the
optimizations discussed in section~\ref{sec:opt}, culmulatively
applied.  The analysis is stopped after consuming 30 minutes of time
or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
use the timeout limits as a lower bound on the actual time required,
thus giving a conservative estimate of improvements.

All benchmarks are calculated as an average of N runs on a [MACHINE
  SPECS].

The benchmarks make use of mutation, first-class control (in the form
of Scheme's {\tt call/cc} operator), compound data such as lists and
vectors, many more primitive forms of data and operations such as
Scheme's numeric tower including exact rationals, complex, and
floating point numbers.

Many benchmarks cause the baseline analyzer to take longer than 30
minutes or to consume more 1 gigabyte of memory, at which point the
analysis is stopped.  This is the case for the largest benchmark
program, which is 3,500 lines of code and takes under a minute in the
most optimized analyzer.  For those benchmarks that did complete on
the baseline, the optimized analyzer outperformed the baseline by a
factor of 475 to 4,382.

@@

\item {\bf matrix} tests whether M is maximal among all matrices of
  the same dimension obtainable by simple reordering of rows and
  columns and negation of any subset of rows and columns.  It is
  written in continuation-passing style (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

@@

\item {\bf earley}:  Earley's parser written by Marc Feeley.

@@

\item {\bf lattice}: enumerates the maps between two lattices (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

@@

\item {\bf graphs}: counts the number of directed graphs with a
  distinguished root and \(k\) vertices, each having out-degree at
  most 2. It makes extensive use of higher-order procedures and is
  written in a continuation-passing style (used by
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
\end{enumerate}

@@

The analysis considered here computes results more like Earl, et al.'s
0-CFA, which times out on the \Church{} benchmark because it does not
widen the store as described for our baseline evaluator.  So even
though it offers a fair point of comparison, a more thorough
evaluation is probably uninformative as the other benchmarks are
likely to timeout as well (and it would require significant effort to
extend their implementation with the features needed to analyze our
benchmark suite).  That implementation is evaluated against much
smaller benchmarks: the largest program is 30 lines.

>>>>

\begin{figure}[t]
\begin{center}
\includegraphics[width=3.2in]{church-relative-space.ps}
\includegraphics[width=3.2in]{church-relative-speed.ps}
\includegraphics[width=3.2in]{church-relative-time.ps}
\vspace{-1.5em}
\end{center}
\caption{Factor improvements over the baseline analyzer for the
  \Church{} benchmark in terms of peak memory usage, the rate of state
  transitions, and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

We have implemented, optimized, and evaluated an analysis framework
supporting higher-order functions, state, first-class control,
compound data, and a large number of primitive kinds of data and
operations such as floating point, complex, and exact rational
arithmetic.  The analysis is evaluated against a suite of benchmarks
drawn from the literature.
%
For each benchmark, we collect analysis times, peak memory usage, and
the rate of states-per-second explored by the analysis for each of the
optimizations discussed in section~\ref{sec:opt}, culmulatively
applied.  The analysis is stopped after consuming 30 minutes of time
or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
use the timeout limits as a lower bound on the actual time required,
thus giving a conservative estimate of improvements.

All benchmarks are calculated as an average of 5 runs, done in
parallel, on an 11-core, 64-bit Intel Xeon machine running at 2.40GHz
with 11Gb of memory.

Many benchmarks cause the baseline analyzer to take longer than 30
minutes or to consume more 1 gigabyte of memory, at which point the
analysis is stopped.  This is the case for the largest benchmark
program, which is 3,500 lines of code and takes under a minute in the
most optimized analyzer.  For those benchmarks that did complete on
the baseline, the optimized analyzer outperformed the baseline by a
factor of two to three orders of magnitude.
% safer to not give exact factors
% 475 to 4,382.

@@

\item {\bf matrix} tests whether a matrix is maximal among all
  matrices of the same dimension obtainable by simple reordering of
  rows and columns and negation of any subset of rows and columns.  It
  is written in continuation-passing style (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

@@

\item {\bf earley}: Earley's parsing algorithm, applied to a 15-symbol
  input according to a simple ambiguous grammar.  A real program,
  applied to small data whose exponential behavior leads to a peak
  heap size of half a gigabyte or more during concrete execution.

@@

\item {\bf lattice}: enumerates the order-preserving maps between two
  finite lattices (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

@@

\item {\bf graphs}: counts the number of directed graphs with a
  distinguished root and \(k\) vertices, each having out-degree at
  most 2. It is written in a continuation-passing style and makes
  extensive use of higher-order procedures---it creates closures
  almost as often as it performs non-tail procedure calls--- (used by
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
\end{enumerate}

@@

The analysis considered here computes results similar Earl, et al.'s
0-CFA implementation, which times out on the \Church{} benchmark
because it does not widen the store as described for our baseline
evaluator.  So even though it offers a fair point of comparison, a
more thorough evaluation is probably uninformative as the other
benchmarks are likely to timeout as well (and it would require
significant effort to extend their implementation with the features
needed to analyze our benchmark suite).  That implementation is
evaluated against much smaller benchmarks: the largest program is 30
lines.
